---
title: "CELLX analysis"
output:
  html_document:
    toc: true
    # toc_float: true
    theme: united
    # theme: cerulean
    # number_sections: true
date: "`r Sys.Date()`"
author: "Mikhail Dozmorov"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r libraries, include=FALSE}
library(knitr)
library(ggplot2)
library(plotly)
```

# Data from CELLX

- Go to [http://54.149.52.246/cgi-bin/RPPA/cellx.cgi](http://54.149.52.246/cgi-bin/RPPA/cellx.cgi)
- Select "Expression" tab out of "CNV"/"Expression"/"Mutation"/"Other" tabs
- Select "RSEM-barplot" option on the sidebar. Read more about RSEM gene expression measures at [https://deweylab.github.io/RSEM/](https://deweylab.github.io/RSEM/)
- Select any cancer-associated gene from [http://cancer.sanger.ac.uk/census](http://cancer.sanger.ac.uk/census), e.g., "ERBB2". Alternatively, use any gene name you may find biologically interesting
- Enter lower-case gene name into the "HUGO" textbox at the bottom of the page
- Click "Submit" - the page will refresh in ~20 sec
- Save the tab-separated data using "Download table" link into `data/RSEM_expression_GENESYMBOL.data.tsv`
- Import the downloaded data into R

```{r}
gene <- "RORC" # Change 
```

# Gene `r gene` analysis

```{r data.setup, include=F}
expr <- read.table(paste0("data/RSEM_expression_", gene, ".data.tsv"), header = TRUE, stringsAsFactors = FALSE)
# expr <- read.table("/Users/mdozmorov/Documents/Work/VCU_work/Paula_Bos/TCGAsurvival/FOXP3/RSEM_expression_FOXP3.data.tsv", header = TRUE, stringsAsFactors = FALSE)

samples <- unique(expr$affy_source)
cancer <- sort(samples[substr(samples, (nchar(samples) - 1), nchar(samples)) != "_N"])
normal <- sort(samples[substr(samples, (nchar(samples) - 1), nchar(samples)) == "_N"])
expr$tissue <- factor(NA, levels = c("Cancer", "Normal"))
expr$tissue[expr$affy_source %in% cancer == T] <- "Cancer"
expr$tissue[expr$affy_source %in% cancer == F] <- "Normal"
expr <- expr[ grep("CCLE", expr$affy_source, invert = TRUE), ]
```    

There are `r dim(expr)[1]` unique samples from `r length(samples)` tissue sources (N~cancer~ = `r length(cancer)`, N~normal~ = `r length(normal)`) indexing expression of the gene *`r names(expr)[3]`*. The graph below shows the expression in of *`r names(expr)[3]`* in normal versus cancerous tissues.    

# Cancer vs. normal expression boxplots, all cancers

```{r expression.plot}
rsem <- ggplot(expr, aes(affy_source, eval(parse(text = gene)))) + 
  geom_boxplot(aes(fill = tissue), outlier.shape = 1) + 
  scale_fill_manual(name = "", values = c("red", "green")) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), panel.grid.major = element_blank()) + 
  labs(title = paste("RSEM Expression", names(expr)[3]), x = NULL) +
  ylab("log2-transformed expression")

rsem
```    

# Cancer vs. normal t-test table

The table below displays differential expression statistics comparing the expression levels of *`r names(expr)[3]`* in normal versus cancerous tissues (where normal comparison tissues are available). See abbreviations of cancer types at [http://www.liuzlab.org/TCGA2STAT/CancerDataChecklist.pdf](http://www.liuzlab.org/TCGA2STAT/CancerDataChecklist.pdf)   

```{r expression.t.table}
# # Check if all normal tissues have a cancer tissue counterpart and vice versa
# table(substr(normal,1,(nchar(normal)-2)) %in% cancer)
# table(cancer %in% substr(normal,1,(nchar(normal)-2)))

# Set up object for storing t-test results
t.table <- data.frame(Cancer_name = NA, 
                      Number_of_normal = NA, 
                      Number_of_cancer = NA, 
                      Mean_expression_normal = NA, 
                      Mean_expression_cancer = NA, 
                      Fold_change = NA,
                      t_statistic = NA, 
                      Degrees_of_freedom = NA, 
                      p_value = NA, 
                      Confidence_interval = NA, 
                      Method = NA)

# Loop over all cancer tissue types to extract information
for (i in 1:length(cancer)) {
    
    # All tissue types have a cancer sample. Fill in cancer tissue info first.
    cname <- substr(cancer[i], 1, (nchar(cancer[i]) - 5))
    if (grepl("TCGA", cancer[i])) {
        cname <- substr(cname, 6, nchar(cname))
    }
    t.table[i, 1] <- cname
    t.table[i, c(3, 5)] <- c(length(which(expr$affy_source == cancer[i])), round(mean(expr[which(expr$affy_source == cancer[i]), 3]), 4))
    
    # Skip other variables if no normal tissues are available for comparison
    if (cancer[i] %in% substr(normal, 1, (nchar(normal) - 2)) == F) {
        t.table$Number_of_normal[i] <- 0
        t.table$Method[i] <- "Not Conducted"
        next
    }
    
    # Otherwise fill in normal tissue info and run t test
    normal_i <- paste0(cancer[i], "_N")
    t.table[i, c(2, 4)] <- c(length(which(expr$affy_source == normal_i)), round(mean(expr[which(expr$affy_source == normal_i), 3]), 4))
    # Add fold change
    t.table$Fold_change[i] <- signif( 2^(t.table$Mean_expression_cancer[i] - t.table$Mean_expression_normal[i]) )

    # Skip t.test errors when N of normal samples is too small; else run and save output
    if (is.null(tryCatch(res <- t.test(expr[which(expr$affy_source == normal_i), 3], expr[which(expr$affy_source ==  cancer[i]), 3], alternative = "two.sided"), error = function(e) NULL))) {
        t.table$Method[i] <- "Not Conducted"
        next
    }
    res <- t.test(expr[which(expr$affy_source == cancer[i]), 3], expr[which(expr$affy_source == normal_i), 3], alternative = "two.sided")
    
    # Write results to t.table
    t.table[i, 7:9] <- c(round(res$statistic, 3), round(res$parameter, 3), round(res$p.value, 6))
    t.table[i, 10:11] <- c(paste0("(", round(res$conf.int[1], 2), ", ", round(res$conf.int[2], 2), ")"), 
        res$method)
}

cancers <- openxlsx::read.xlsx("data.TCGA/TCGA_cancers.xlsx")

t.table <- t.table[ !is.na(t.table$t_statistic), ]
t.table <- left_join(cancers, t.table, by = c("Acronym" = "Cancer_name"))
DT::datatable(t.table[ order(t.table$p_value), ], width = 800)
```    

# Cancer vs. normal expression boxplots, individual cancers

The figure below is an interactive plot that highlights the the differences in *`r names(expr)[3]`* expression distribution in normal and cancerous tissues for each type of cancer. 

```{r interactive.plot, warning=F, fig.height=12}
# Add t test info to dataset to display means and test results in the graph
expr$Cancer <- substr(expr$affy_source, 1, (nchar(expr$affy_source) - 5))
expr$Cancer[expr$tissue == "Normal"] <- substr(expr$affy_source[expr$tissue == "Normal"], 1, (nchar(expr$affy_source[expr$tissue == "Normal"]) - 7))
expr$Cancer[grepl("TCGA", expr$Cancer)] <- substr(expr$Cancer[grepl("TCGA", expr$Cancer)], 6, nchar(expr$Cancer[grepl("TCGA", expr$Cancer)]))

# length(unique(expr$Cancer)) 
# table(expr$Cancer)
expr2 <- merge(expr, t.table, by.x = "Cancer", by.y = "Acronym")
expr2 <- expr2[!is.na(expr2$Mean_expression_normal), ]

## Boxplot version: looks nice but the hover text doesn't work for displaying t.test statistics
rsem_int <- ggplot(expr2, aes(tissue, eval(parse(text = gene)), text = paste("p:", p_value))) + 
  geom_boxplot(aes(fill = tissue, text = paste("p:", p_value)), outlier.shape = 1) + 
  scale_fill_manual(name = "", values = c("red", "green")) + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), panel.grid.major = element_blank()) + 
  labs(title = paste("RSEM Expression", names(expr)[3], "\n"), x = NULL) + 
  facet_wrap(~Cancer, nrow = 4) + guides(fill = F)  +
  ylab("log2-transformed expression")

ggplotly(rsem_int, tooltip = c("colour", "text"))
```    
<!--
# Cancer vs. normal expression~significance plot

The figure below is an interactive plot that highlights the relationship between *`r names(expr)[3]`* expression in normal and cancerous tissues for each type of cancer, colored by the significance of the differences.

```{r interactive.coord.plot}
# Coordinate plot version
t.int <- ggplot(t.table[complete.cases(t.table), ], aes(Mean_expression_normal, Mean_expression_cancer)) + 
    geom_point(aes(text = paste("Name:", Acronym, ", t:", t_statistic), color = p_value)) + labs(title = paste("Expression of", names(expr)[3], "Across Cancer Types"), x = "\nExpression Level in Normal Tissue", y = "Expression Level in Cancer Tissue\n") + 
    scale_color_gradient(low = "red", high = "black")

ggplotly(t.int)
```    
-->

