---
title: "Correlations between two genes across all cancers"
output:
  html_document:
    toc: true
    # toc_float: true
    theme: united
    # theme: cerulean
    # number_sections: true
date: "`r Sys.Date()`"
author: "Mikhail Dozmorov"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r libraries, include=FALSE}
library(openxlsx)
library(MDmisc)
library(org.Hs.eg.db)
library(KEGG.db)
library(TCGA2STAT)
library(ggplot2)
library(reshape2)
# library(clusterProfiler)
library(pathview)
library(annotables)
# Remove non-canonical chromosome names
grch38 <- grch38[ !(grepl("_", grch38$chr) | grepl("GL", grch38$chr)), ]
grch38 <- grch38[, c("symbol", "description")]
grch38 <- grch38[ complete.cases(grch38) , ]
grch38 <- grch38[ !duplicated(grch38), ]
```

```{r functions}
# A function to load TCGA data, from remote repository, or a local R object
load_data <- function(disease = cancer, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE) {
  FILE = paste0(data_dir, "/mtx_", disease, "_", data.type, "_", type, ".rda") # R object with data
  if (all(file.exists(FILE), !(force_reload))) {
    # If the data has been previously saved, load it
    load(file = FILE)
  } else {
    # If no saved data exists, get it from the remote source
    mtx <- getTCGA(disease = disease, data.type = data.type, type = type, clinical = TRUE)
    save(file = FILE, list = c("mtx")) # Save it
  }
  return(mtx)
}
# A wrapper function to perform all functional enrichment analyses.
# Helper function to save non-empty results
save_res <- function(res, fileName = fileName, wb = wb, sheetName = "KEGG") {
  if (nrow(res) > 0) {
    openxlsx::addWorksheet(wb = wb, sheetName = sheetName)
    openxlsx::writeData(wb, res, sheet = sheetName)
    openxlsx::saveWorkbook(wb, fileName, overwrite = TRUE)
  }
}

# A wrapper to save the results
save_enrichr <- function(up.genes = up.genes, dn.genes = NULL, databases = "KEGG_2016", fdr.cutoff = 1, fileNameOut = NULL, wb = NULL) {
  print(paste("Running", databases, "analysis", sep = " "))
  if (is.null(dn.genes)) {
    res.kegg <- enrichGeneList(up.genes, databases = databases, fdr.cutoff = 1)
  } else {
    res.kegg <- enrichFullGeneList(up.genes, dn.genes, databases = databases, fdr.cutoff = 1)
  }
  
  res.kegg$pval <- formatC(res.kegg$pval, digits = 3, format = "e")
  res.kegg$qval <- formatC(res.kegg$qval, digits = 3, format = "e")
  if (!is.null(fileNameOut)) {
    if (nchar(databases) > 30) databases <- paste0(substr(databases, 1, 20), "_", substr(databases, nchar(databases) - 8, nchar(databases))) # If a database is longer that 30 characters, keep first 20 and last 10 characters
    save_res(res.kegg, fileNameOut, wb = wb, sheetName = databases)
  }
  # Pause for a few seconds
  pause_sec <- round(runif(1, min = 1, max = 10))
  Sys.sleep(pause_sec)
  return(res.kegg)
}
```

```{r settings}
system("mkdir -p data")
system("mkdir -p results")
# Path where the downloaded data is stored
data_dir = "/Users/mdozmorov/Documents/Data/GenomeRunner/TCGAsurvival/data" # Mac
# data_dir = "F:/Data/GenomeRunner/TCGAsurvival/data" # Windows

# Selected genes, TWO genes only
selected_genes <- c("ANLN", "VIM")
# Check if the names exist
genes <- readLines("data.TCGA/TCGA_genes.txt")
setdiff(selected_genes, genes) # Should be 0, meaning selected genes are among the TCGA genes

data.type = "RNASeq2"; type = "" 
# All cancers with RNASeq2 data
cancer_RNASeq2 = c("ACC", "BLCA", "BRCA" , "CESC", "CHOL", "COAD", "COADREAD", "DLBC", "ESCA", "GBM", "GBMLGG", "HNSC", "KICH", "KIPAN", "KIRC", "KIRP", "LGG", "LIHC", "LUAD", "LUSC", "MESO", "OV", "PAAD", "PCPG", "PRAD", "READ", "SARC", "SKCM", "STAD", "TGCT", "THCA", "THYM", "UCEC", "UCS")
# fileNameIn <- (paste0("data/All_expression_", data.type, "_", type, ".Rda")) # Save expression data
# fileNameOut <- paste0("results/All_correlation_", selected_genes, "_", data.type, "_", type, ".Rda") # Save correlation data
# fileNameRes <- paste0("results/All_results_", selected_genes, "_", data.type, "_", type, ".xlsx") # Save results
# Or, one cancer
# cancer_RNASeq2 = c("LUAD")
fileNameIn <- (paste0("data/", cancer_RNASeq2, "_expression_", data.type, "_", type, ".Rda"))  # Save expression data
fileNameOut <- paste0("results/", cancer_RNASeq2, "_correlation_", selected_genes, "_", data.type, "_", type, ".Rda") # Save correlation data
fileNameRes <- paste0("results/", cancer_RNASeq2, "_results_", selected_genes, "_", data.type, "_", type, ".xlsx") # Save results

# Correlation type
corr_type    <- "pearson"
# Correlation cutoffs
corr_cutoff  <- 0.2
pval_cutoff  <- 0.05
# Enrichment cutoffs
p.adj.cutoff <- 0.1
fdr.cutoff   <- 0.3
```

- Each cancer is processed separately. Headers "Processing cancer XXX" start results section for a given cancer.
- Linear regression results between expression of two genes are outputted. "Adj R2" and "P" - percent variability explained by regression model (higher the better) and its p-value (smaller the better).
- Correlation graph shows expression of one gene plotted vs. the other gene in a given cancer, with linear regression line and confidence intervals around it.
- Expression graph shows boxplots of expression of two genes in a given cancer. If one, or both, genes are low expressed, any correlation/regression results should be interpreted with caution.

```{r correlations, fig.height=3, fig.width=3}
  all_corrs <- list() # List to store cancer-specific correlationa
  all_pvals <- list() # List to store cancer-specific p-values
  # Get correlation matrixes for the gene of interest in each cancer
  for (cancer_type in cancer_RNASeq2) {
    pandoc.header(paste0("Processing cancer ", cancer_type))
    # Prepare expression data
    mtx <- load_data(disease = cancer_type, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
    expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
    # Attach column with gene symbols
    expr <- data.frame(hgnc = colnames(expr), t(expr))
    # Correlation for a pair of genes
    cancer_rcorr <- Hmisc::rcorr(as.numeric(expr[selected_genes[1], 2:ncol(expr)]), 
                                 as.numeric(expr[selected_genes[2], 2:ncol(expr)]), 
                                 type = corr_type)
    # Matrix to plot correlations
    mtx_to_plot <- data.frame(gene1 = log2(as.numeric(expr[selected_genes[1], 2:ncol(expr)]) + 1),
                              gene2 = log2(as.numeric(expr[selected_genes[2], 2:ncol(expr)]) + 1))
    colnames(mtx_to_plot) <- c(selected_genes[1], selected_genes[2]) # Assign column names
    # Quick regression
    fit <- lm(mtx_to_plot[, 1] ~ mtx_to_plot[, 2])
    # Plot correlation graph
    p <- ggplot(mtx_to_plot, aes(x = eval(parse(text = selected_genes[1])), y = eval(parse(text = selected_genes[2])))) +
      geom_point(shape = 1) +
      geom_smooth(method = lm) +
      xlab(selected_genes[1]) +
      ylab(selected_genes[2]) + 
      ggtitle("Correlation graph")
    # Fit statistics
    print(paste("Adj R2 = ",   signif(summary(fit)$adj.r.squared, 5),
                "Intercept =", signif(fit$coef[[1]],5 ),
                " Slope =",    signif(fit$coef[[2]], 5),
                " P =",        signif(summary(fit)$coef[2,4], 5)))
    plot(p)
    # Plot expression graph
    mtx_to_plot <- melt(mtx_to_plot)
    p <- ggplot(mtx_to_plot, aes(x = variable, y = value, fill = variable)) +
      geom_boxplot() +
      ggtitle("Expression plot") +
      xlab("Gene") + ylab("log2 expression")
    plot(p)
    # Store the results
    all_corrs <- c(all_corrs, list(cancer_rcorr$r[1, 2]))
    all_pvals <- c(all_pvals, list(cancer_rcorr$P[1, 2]))
    names(all_corrs)[length(all_corrs)] <- names(all_pvals)[length(all_pvals)] <- cancer_type
  }
```

# Correlation (`r corr_type`) between `r selected_genes[1]` and `r selected_genes[2]`

- "corr", "pval" - `r corr_type` correlation coefficient and the p-value of correlations
- Sort the table by "corr", "pval" columns

```{r}
res_corr <- data.frame(Acronym = cancer_RNASeq2, corr = unlist(all_corrs), pval = unlist(all_pvals))
cancers  <- openxlsx::read.xlsx("data.TCGA/TCGA_cancers.xlsx")
res_corr <- left_join(res_corr, cancers, by = c("Acronym"))
res_corr <- round_df(res_corr)
res_corr <- res_corr[ order(res_corr$corr, decreasing = TRUE), ]
DT::datatable(res_corr, options = list(pageLength = 50))
```

