---
title: "PPI and network analysis"
output:
  html_document:
    toc: true
    # toc_float: true
    theme: united
    # theme: cerulean
    # number_sections: true
date: "`r Sys.Date()`"
author: "Mikhail Dozmorov"
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

## Libraries

```{r}
library(aracne.networks)
library(networkD3)
library(igraph)
```

## Settings

```{r}
selected_genes <- c("SDC1")
max_to_plot <- 60 # Maximum number of PPI links to plot
```

## PPI databases 

- The tables contain PPI interactions for the selected gene `r selected_genes`.
- Search box in the tables to search for for interactions of interest. E.g., search for "cxcl8" to find an interaction of `r selected_genes` with IL-8
- Two networks, if plotted, are the same. One is dynamic (drag nodes, zoom with mouse wheel), another is static.
- If the selected gene `r selected_genes` has more than `r max_to_plot` connections, only the table is outputted, the networks are not plotted.

### `STRING`, http://string-db.org/

STRING v10: protein-protein interaction networks, integrated over the tree of life. https://www.ncbi.nlm.nih.gov/pubmed/25352553

```{r eval = TRUE}
# STRING data
load(file = "/Users/mdozmorov/Documents/Work/VCU_work/Mark/disease-coherence/data/9606.protein.links.v10.rda")

selected <- my_adj_list[ (my_adj_list$from %in% selected_genes) | (my_adj_list$to %in% selected_genes), , drop = FALSE ]
selected <- selected[complete.cases(selected), ]
print(paste0("Number of proteins interacting witn ", selected_genes, ": ", nrow(selected)))
DT::datatable(selected)
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # NetworkD3 plot
  simpleNetwork(selected, Source = "from", Target = "to", fontSize = 10, linkColour = "black", nodeColour = "blue", opacity = 0.8, zoom = TRUE)
}
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # igraph plot
  net <- graph_from_data_frame(d = selected, directed = TRUE)
  plot(net, edge.arrow.size=.6, vertex.label.family = "Arial")
}
```

### `I2D`, http://ophid.utoronto.ca/ophidv2.204/index.jsp

I2D (Interologous Interaction Database) is an on-line database of known and predicted mammalian and eukaryotic protein-protein interactions

```{r}
load(file = "/Users/mdozmorov/Documents/Work/VCU_work/Mark/disease-coherence/data/i2d.2_9.Public.HUMAN.tab.rda")

selected <- my_adj_list[ (my_adj_list$from %in% selected_genes) | (my_adj_list$to %in% selected_genes), , drop = FALSE ]
selected <- selected[complete.cases(selected), ]
print(paste0("Number of proteins interacting witn ", selected_genes, ": ", nrow(selected)))
DT::datatable(selected)
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # NetworkD3 plot
  simpleNetwork(selected, Source = "from", Target = "to", fontSize = 10, linkColour = "black", nodeColour = "blue", opacity = 0.8, zoom = TRUE)
}
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # igraph plot
  net <- graph_from_data_frame(d = selected, directed = TRUE)
  plot(net, edge.arrow.size=.6, vertex.label.family = "Arial")
}
```

### `Biogrid`, https://thebiogrid.org/

BioGRID is an interaction repository with data compiled through comprehensive curation efforts.

```{r eval = TRUE}
load(file = "/Users/mdozmorov/Documents/Work/VCU_work/Mark/disease-coherence/data/biogrid.rda")

selected <- my_adj_list[ (my_adj_list$from %in% selected_genes) | (my_adj_list$to %in% selected_genes), , drop = FALSE ]
selected <- selected[complete.cases(selected), ]
print(paste0("Number of proteins interacting witn ", selected_genes, ": ", nrow(selected)))
DT::datatable(selected)
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # NetworkD3 plot
  simpleNetwork(selected, Source = "from", Target = "to", fontSize = 10, linkColour = "black", nodeColour = "blue", opacity = 0.8, zoom = TRUE)
}
```
```{r}
if (nrow(selected) <= max_to_plot) {
  # igraph plot
  net <- graph_from_data_frame(d = selected, directed = TRUE)
  plot(net, edge.arrow.size=.6, vertex.label.family = "Arial")
}
```

## Other resources

### `BioPlex`, http://bioplex.hms.harvard.edu/index.php
The BioPlex (biophysical interactions of ORFeome-based complexes) network is the result of creating thousands of cell lines with each expressing a tagged version of a protein from the ORFeome collection. Immunopurification of the tagged protein and detection of associated proteins by mass spectrometry are the building blocks of the network. 

### `IntAct`, https://www.ebi.ac.uk/intact/
IntAct Molecular Interaction Database. IntAct provides a freely available, open source database system and analysis tools for molecular interaction data. All interactions are derived from literature curation or direct user submissions and are freely available.

### `InBio`, https://www.intomics.com/inbio/map/#home
InBio Map is a high coverage, high quality, convenient and transparent platform for investigating and visualizing protein-protein interactions. Visualization: http://apps.broadinstitute.org/genets#InWeb_InBiomap

### `HIPPIE`, http://cbdm-01.zdv.uni-mainz.de/~mschaefer/hippie/index.php
Human Integrated Protein-Protein Interaction rEference

### `HuRI`, http://interactome.baderlab.org/
Human Reference Protein Interactome Project

One of the long-term goals at the Center for Cancer Systems Biology (CCSB) is to generate a first reference map of the human protein-protein interactome network. To reach this target, we are identifying binary protein-protein interactions (PPIs) by systematically interrogating all pairwise combinations of predicted human protein-coding genes using proteome-scale technologies. Our approach to map high-quality PPIs is based on using yeast two-hybrid (Y2H) as the primary screening method followed by validation of subsets of PPIs in multiple orthogonal assays for binary PPI detection.

### `PINA`, http://omics.bjcancer.org/pina/home.do
Protein Interaction Network Analysis (PINA), which collected and annotated six other public PPI databases (MINT, IntAct, DIP, BioGRID, HPRD, and MIPS/MPact).

