---
title: "TCGA BRCA survival analysis"
output:
  html_document:
    toc: true
    # toc_float: true
    theme: united
    # theme: cerulean
    # number_sections: true
date: "`r Sys.Date()`"
author: "Mikhail Dozmorov"
bibliography: data.TCGA/TCGA.bib
csl: styles.ref/genomebiology.csl
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is', fig.width = 10, fig.height = 6) #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

# Methods

## Survival analysis of gene expression data from TCGA

Level 3 gene expression data summarized as RSEM values was obtained using the `TCGA2STAT` R package v 1.2, along with the corresponding clinical annotations. Data for each of the 34 cancers was obtained separately. The data was log2-transformed and analyzed using Kaplan-Meyer curves and Cox proportional hazard model. Each gene of interest was analyzed for its effect on survival by separating patients into high/low expression subgroups. A modified approach from [@Mihaly:2013aa] was used to estimate the best gene expression cutoff that separates high/low expression subgroups with differential survival.

We took the advantage of the availability of clinical annotations. To identify if expression of a gene of interest affects survival in any specific clinical subgroup, subsets of patients annotated with specific clinical annotations were selected (e.g., “males” or “females” in the “gender” clinical annotation). Subgroups with < 40 patients were not considered.

## Differential expression analysis

Samples in the selected cancer cohort were sorted by expression of the selected genes. Differentially expressed genes were detected between samples in the upper 75 percentile of the expression gradient and samples in the lower 25 percentile using `limma` v 3.32.6 R package [@Ritchie:2015aa; @Smyth:2004aa]. P-values were corrected for multiple testing using False Discovery Rate (FDR) method [@Benjamini:1995aa]. Genes differentially expressed at FDR < 0.01 were selected for further analysis.


# Libraries

```{r libraries}
# library(TCGA2STAT)
library(ggplot2)
library(reshape2)
if (!require(survplot)) {
  install.packages("misc/survplot_0.0.7.tar.gz", repos = NULL, type="source")
}
source("Supplemental_R_script_1.R")
```

# Helper functions

```{r functions}
# A function to load TCGA data, from remote repository, or a local R object
load_data <- function(disease = cancer, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE) {
  FILE = paste0(data_dir, "/mtx_", disease, "_", data.type, "_", type, ".rda") # R object with data
  if (all(file.exists(FILE), !(force_reload))) {
    # If the data has been previously saved, load it
    load(file = FILE)
  } else {
    # If no saved data exists, get it from the remote source
    mtx <- getTCGA(disease = disease, data.type = data.type, type = type, clinical = TRUE)
    save(file = FILE, list = c("mtx")) # Save it
  }
  return(mtx)
}

# A function to get data overview
summarize_data <- function(mtx = mtx) {
  print(paste0("Dimensions of expression matrix, genex X patients: ", paste(dim(mtx$dat), collapse = " ")))
  print(paste0("Dimensions of clinical matrix, patients X parameters: ", paste(dim(mtx$clinical), collapse = " ")))
  print(paste0("Dimensions of merged matrix, patients X parameters + genes: ", paste(dim(mtx$merged.dat), collapse = " ")))
  print("Head of the merged matrix")
  print(mtx$merged.dat[1:5, 1:10])
  print("Head of the clinical matrix")
  print(mtx$clinical[1:5, 1:7])
  print("List of clinical values, and frequency of each variable: ")
  clin_vars <- apply(mtx$clinical, 2, function(x) length(table(x[ !(is.na(x) & x != "" )]))) %>% as.data.frame()
  # Filter clinical variables to have at least 2, but no more than 10 categories,
  # And they are not dates
  clin_vars <- clin_vars[ as.numeric(clin_vars$.) > 1 & as.numeric(clin_vars$.) < 10 & !grepl("years|days|date|vital|OS|RFS|TIME|sample_type", rownames(clin_vars), perl = TRUE) , , drop = FALSE]
  print(kable(clin_vars))
  return(rownames(clin_vars))
}

# A function to create expression matrix
make_expression_matrix <- function(mtx = mtx, disease = cancer, data.type = data.type, type = type, results_dir = results_dir) {
  # transposed expression matrix (genes start at column 4) 
  mtx.expression <- mtx$merged.dat[, 4:ncol(mtx$merged.dat) ] %>% t 
  # Set column names as patient IDs
  colnames(mtx.expression) <- mtx$merged.dat$bcr 
  # Set row names as probe IDs
  rownames(mtx.expression) <- colnames(mtx$merged.dat)[ 4:ncol(mtx$merged.dat) ] 
  # Save gzipped matrix
  fileName.gz <- gzfile(paste0(results_dir, "/mtx_", disease, "_", data.type, "_", type, "_1expression.txt.gz"), "w")
  write.table(mtx.expression, fileName.gz, sep = ";", quote = FALSE)
  close(fileName.gz)
}
```

# Settings

```{r settings}
# Path where the downloaded data is stored
data_dir = "/Users/mdozmorov/Documents/Data/GenomeRunner/TCGAsurvival/data" # Mac
# data_dir = "/Users/mdozmorov/Documents/nobackup" # Mac
# data_dir = "F:/Data/GenomeRunner/TCGAsurvival/data" # Windows
# data_dir = "D:"
# results_dir = "/Users/mdozmorov/Dropbox" # Path where the results are stored

# Cancer types: http://www.liuzlab.org/TCGA2STAT/CancerDataChecklist.pdf
# Data types: http://www.liuzlab.org/TCGA2STAT/DataPlatforms.pdf
# Expression types: http://www.liuzlab.org/TCGA2STAT/DataPlatforms.pdf
# Clinical values: http://www.liuzlab.org/TCGA2STAT/ClinicalVariables.pdf

# General settings
data.type = "RNASeq2"
type = "" 
# data.type = "miRNASeq" # miRNASeq - "count" for raw read counts (default); "rpmmm" for normalized read counts
# type = "rpmmm"
# data.type = "Mutation"
# type = "somatic"
# type="all"
# data.type = "Methylation"
# type = "450K"
# clinical = TRUE

# Breast cancer
cancer = "BRCA" 
# # Ovarian cancer
# cancer = "OV"
# # Liver hepatocellular carcinoma
# cancer = "LIHC" 
# # Head and Neck squamous cell carcinoma
# cancer = "HNSC" 
# # Sarcoma
# cancer = "SARC" 
# # Pancreatic cancer
# cancer = "PAAD" 
# # Lung adenocarcinoma
# cancer = "LUAD"
# # Lung squamous cell carcinoma
# cancer = "LUSC"
```

# Load data

```{r loadData}
mtx <- load_data(disease = cancer, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
clinical_annotations <- summarize_data(mtx = mtx)
```

```{r processData}
# Prepare expression data
expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
# Filter out low expressed genes
# Should be more than 90% of non-zero values
ff <- genefilter::pOverA(p = 0.9, A = 0, na.rm = TRUE) 
expr <- expr[, apply(expr, 2, ff)] 
expr <- data.frame(AffyID = mtx$merged.dat$bcr, expr, stringsAsFactors = FALSE)

# Prepare clinical data
clin <- mtx$merged.dat[, 1:3]
colnames(clin)[1] <- "AffyID"
```

# Select genes

```{r selectedGenes}
# Select genes of interest
# selected_genes = c("TMEM219", "IGFBP3") # BRCA
# selected_genes = c("BCL2L11", "MYC", "MYCN") # LUAD
# selected_genes = c("LAPTM4B", "PIP5K1C") # HNSC
# selected_genes = c("MTDH", "SND1", "TAF2") # LIHC
# selected_genes = c("NF1") # LIHC
# selected_genes = c("ANLN") # PAN
# selected_genes = c("TAF2") # LIHC
# selected_genes = c("CPEB2") # BRCA
# selected_genes = c("PECAM1", "S1PR1", "SPNS2", "TEK", "TIE1"); cancer_RNASeq2 = c("PAAD")
# selected_genes = c("hsa.mir.142"); cancer_RNASeq2 = c("BRCA")
selected_genes = "MIA"; cancer_RNASeq2 = c("BRCA")
# selected_genes = "DYRK1A"
# selected_genes = "DCAF7"
# selected_genes = "MBD2" # AML
# selected_genes = c("FOXP3") # BRCA
# selected_genes = c("TNFRSF1B") # TNFR2, BRCA
# selected_genes = c("RORC") # BRCA
```

### Analysis 1: Selected genes, selected cancers, no clinical annotations

```{r analysis1}
system("mkdir res") # Create results folder
kmplot(expr, clin, event_index=2, time_index=3,  affyid = selected_genes, auto_cutoff="true", transform_to_log2 = TRUE, cancer_type = cancer, fileType = "png", use_survminer = FALSE)
# Rename the results folder
system("mv res res.genes.Analysis1")
```

```{r analysisExploratory, eval = FALSE}
### Exploratory: All genes, selected cancers, no clinical annotations
# kmplot(expr, clin, event_index=2, time_index=3,  affyid = "", auto_cutoff="true", transform_to_log2 = TRUE)
```

### Analysis 2: Selected genes, all (or selected) cancers, no clinical annotations

```{r analysis2}
# All cancers with RNASeq2 data
cancer_RNASeq2 = c("ACC", "BLCA", "BRCA", "CESC", "CHOL", "COAD", "COADREAD", "DLBC", "ESCA", "GBM", "HNSC", "KICH", "KIPAN", "KIRC", "KIRP", "LIHC", "LUAD", "LUSC", "MESO", "OV", "PAAD", "PCPG", "PRAD", "READ", "SARC", "SKCM", "STAD", "TGCT", "THCA", "THYM", "UCEC", "UCS") # "GBMLGG", "LGG", 
data.type = "RNASeq2"; type = "" 
system("mkdir res") # Create results folder
for (cancer_type in cancer_RNASeq2) {
  print(paste0("Processing cancer ", cancer_type))
  mtx <- load_data(disease = cancer_type, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
  # Prepare expression data
  expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
  # Filter out low expressed genes
  # Should be more than 90% of non-zero values
  # ff <- genefilter::pOverA(p = 0.9, A = 0, na.rm = TRUE) 
  # expr <- expr[, apply(expr, 2, ff)] 
  expr <- data.frame(AffyID = mtx$merged.dat$bcr, expr, stringsAsFactors = FALSE)
  # Prepare clinical data
  clin <- mtx$merged.dat[, 1:3]
  colnames(clin)[1] <- "AffyID"
  # Run survival analysis for selected genes
  kmplot(expr, clin, event_index=2, time_index=3,  affyid = selected_genes, auto_cutoff="true", transform_to_log2 = TRUE, cancer_type = cancer_type, fileType = "png", use_survminer = FALSE)
}
### Plot the results of one gene across all cancers
library(dplyr)
library(ggplot2)
# Read in analysis natrix
mtx <- read.table("res/global_stats.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE, fill = TRUE)
# Add -log10-transformed p-value
mtx <- mtx %>% mutate(log10.pval = -1 * log10(p.value))
# Print into PDF
png(paste0("res/", selected_genes, "_all_TCGA_cancers.png"))
mtx %>% subset(Gene == selected_genes) %>% 
  ggplot(aes(x = factor(Cancer, levels(factor(Cancer))[order(log10.pval)]), y = log10.pval)) + 
  geom_bar(stat = "identity") + 
  theme(legend.position="none") +
  labs(x="Cancer", y="-log10(p-value)") +
  coord_flip()
dev.off()
# Rename the results folder
system("mv res res.genes.Analysis2")
```

### Analysis 3: Selected genes, all (or, selected) cancers, all unique categories

```{r analysis3}
cancer_RNASeq2 = c("BRCA")
system("mkdir res") # Create results folder
for (cancer_type in cancer_RNASeq2) {
  print(paste0("Processing cancer ", cancer_type))
  mtx <- load_data(disease = cancer_type, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
  # BRCA-specific - replace original annotations with XENA
  mtx$clinical <- read.csv("data.TCGA/BRCA_XENA_clinical.csv", row.names = 1)
  
  clinical_annotations <- summarize_data(mtx = mtx)
  # Prepare expression data
  expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
  # Filter out low expressed genes
  # Should be more than 90% of non-zero values
  # ff <- genefilter::pOverA(p = 0.9, A = 0, na.rm = TRUE) 
  # expr <- expr[, apply(expr, 2, ff)] 
  expr <- data.frame(AffyID = mtx$merged.dat$bcr, expr, stringsAsFactors = FALSE)
  # Prepare clinical data
  clin <- mtx$merged.dat[, 1:3]
  colnames(clin)[1] <- "AffyID"
  # Full clinical information
  clin_full <- mtx$clinical
  # Match to the order of small clinical annitation
  clin_full <- clin_full[rownames(clin_full) %in% clin$AffyID, ]
  clin_full <- clin_full[match(clin$AffyID, rownames(clin_full)), ]

  # For each clinical annotation
  for (annotation in clinical_annotations) { 
    # Get the number of patients per category in the current annotation
    annotations <- table(clin_full[, annotation], useNA = "no") 
    # How many categories in the current annotation
    num_of_annot_categories <- length(annotations) 
    # How many categories to select at one time
    for (num_of_selected_categories in 1:num_of_annot_categories) {
      # Select patients annotated with this categories
      patients <- rownames(clin_full)[ clin_full[, annotation] %in% names(annotations)[ num_of_selected_categories] ]
      # Get their index in the clin and expr matrixes
      index_patients <- which(clin$AffyID %in% patients)
      # If the number of patients annotated with the combination of categories is large enough, proceed
      if (length(index_patients) > 40) {
        print(paste("Processing annotation:", annotation, 
                    ", categories:", names(annotations)[ num_of_selected_categories],
                    ", number of patients:", length(index_patients)))
        # Get a subset of clinical information for these patients
        clin_selected <- clin[ index_patients, ]
        # Get a subset of expression information for these patients
        expr_selected <- expr[ index_patients, ]
        # For this subset of expression, filter out low expressed genes
        index_genes <- apply(expr_selected %>% dplyr::select(-AffyID), 2, ff) # index of expression values to keep
        expr_selected <- cbind(expr_selected$AffyID, dplyr::select(expr_selected, -AffyID)[, index_genes]) # patient IDs and expression values to keep
        # Perform actual survival analysis
        kmplot(expr_selected, clin_selected, event_index=2, time_index=3,  affyid = selected_genes, auto_cutoff="true", transform_to_log2 = TRUE, cancer_type = paste(c(cancer_type, annotation, names(annotations)[ num_of_selected_categories] ), collapse = "-"), fileType = "png", use_survminer = FALSE)
      }
    }
  }
}
# Rename the results folder
system("mv res res.genes.Analysis3")
```

```{r analysis4, eval = FALSE}
### Analysis 4: Selected genes, selected cancers, all combinations of clinical annotations
system("mkdir res") # Create results folder
clin_full <- mtx$clinical # Full clinical information
# For each clinical annotation
for (annotation in clinical_annotations) { 
  # Get the number of patients per category in the current annotation
  annotations <- table(clin_full[, annotation], useNA = "no") 
  # How many categories in the current annotation
  num_of_annot_categories <- length(annotations) 
  # How many categories to select at one time
  for (num_of_selected_categories in 1:num_of_annot_categories) {
    # All combinations of categories, m categories selected at one time
    combination_of_categories <- combn(x = names(annotations), m = num_of_selected_categories)
    # For each combination of categories (column)
    for (combination in 1:ncol(combination_of_categories)) {
      # Select patients annotated with this combination of categories
      patients <- rownames(clin_full)[ clin_full[, annotation] %in% combination_of_categories[, combination]]
      # Get their index in the clin and expr matrixes
      index_patients <- which(clin$AffyID %in% patients)
      # If the number of patients annotated with the combination of categories is large enough, proceed
      if (length(index_patients) > 40) {
        print(paste("Processing annotation:", annotation, 
                    ", categories:", paste(combination_of_categories[, combination], collapse = ","),
                    ", number of patients:", length(index_patients)))
        # Get a subset of clinical information for these patients
        clin_selected <- clin[ index_patients, ]
        # Get a subset of expression information for these patients
        expr_selected <- expr[ index_patients, ]
        # For this subset of expression, filter out low expressed genes
        index_genes <- apply(expr_selected %>% dplyr::select(-AffyID), 2, ff) # index of expression values to keep
        expr_selected <- cbind(expr_selected$AffyID, select(expr_selected, -AffyID)[, index_genes]) # patient IDs and expression values to keep
        # Perform actual survival analysis
        kmplot(expr_selected, clin_selected, event_index=2, time_index=3,  affyid = selected_genes, auto_cutoff="true", transform_to_log2 = TRUE, cancer_type = paste(c(cancer, annotation, combination_of_categories[, combination]), collapse = "-"), fileType = "png", use_survminer = FALSE)
      }
    }
  }
}
# Rename the results folder
system("mv res res.genes.Analysis4")
```

```{r}
# Analysis 5: Clinical-centric analysis. Selected cancer, selected clinical subcategory, survival difference between all pairs of subcategories
system("mkdir -p res")
# Select cancer and load expression data
cancer <- "BRCA"
mtx <- load_data(disease = cancer, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)

# Prepare clinical data
mtx$clinical <- read.csv("data.TCGA/BRCA_XENA_clinical.csv", row.names = 1)
clin <- mtx$merged.dat[, 1:3]
colnames(clin)[1] <- "AffyID"
# Full clinical information
clin_full <- mtx$clinical
# Match to the order of small clinical annitation
clin_full <- clin_full[rownames(clin_full) %in% clin$AffyID, ]
clin_full <- clin_full[match(clin$AffyID, rownames(clin_full)), ]
# Select clinical category and subcategories
clinical_annotations <- summarize_data(mtx = mtx)
clinical_annotations_selected <- "PAM50Call_RNAseq" # For PAM50 classifier
clin <- cbind(clin, clin_full[, clinical_annotations_selected, drop = FALSE])

print(paste0("Number of patients in each subcategory, in the ", clinical_annotations_selected, " category"))
table(clin[, clinical_annotations_selected]) 
# All pairs of subcategories
group_combinations <- combn(unique(clin[!is.na(clin[, clinical_annotations_selected]), clinical_annotations_selected]), 2) 
# Survival analysis for all group combinations
for (pair in 1:ncol(group_combinations)) {
  # KM analysis on subcategories
  kmplot.clin(clin = clin, event_index=2, time_index=3, clinical_annotations = clinical_annotations_selected, group1 = group_combinations[1, pair], group2 = group_combinations[2, pair], cancer_type = cancer, fileType = "png", use_survminer = FALSE)
}


# Plot expression boxplots for the selected gene in selected subcategories
# selected_genes = c("SDC1") # BRCA
# Prepare expression data
expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
# Filter out low expressed genes
# Should be more than 90% of non-zero values
# ff <- genefilter::pOverA(p = 0.9, A = 0, na.rm = TRUE) 
# expr <- expr[, apply(expr, 2, ff)] 
expr <- data.frame(AffyID = mtx$merged.dat$bcr, expr, stringsAsFactors = FALSE)
expr <- expr[ expr$AffyID %in% clin$AffyID, ]
expr <- expr[match(clin$AffyID, expr$AffyID), ]
# A matrix to plot
mtx_to_plot <- data.frame(Gene = log2(expr[, selected_genes] + 1), Clinical = clin[, clinical_annotations_selected])
mtx_to_plot <- mtx_to_plot[complete.cases(mtx_to_plot), ]
save(mtx_to_plot, file = "res/mtx_to_plot.rda")
# Plot and save
p <- ggplot(melt(mtx_to_plot, id.vars = "Clinical"), aes(x = Clinical, y = value, fill = Clinical)) +
  geom_boxplot() +
  ylab("log2 expression")
plot(p)
ggsave(filename = paste0("res/", cancer, "_", selected_genes, "_", clinical_annotations_selected, ".png"), p, width = 5, height = 4, device = "png")
system("mv res res.genes.Analysis5")
```

```{r eval = FALSE}
## Analysis 6: Dimensionality reduction of a gene signature across all cancers using NMF, PCA, or FA
# For each cancer, extracts gene expression of a signature, reduces its dimensionality,
# plots a heatmap sorted by the first component, biplots, saves eigenvectors in files
# named after cancer, signature, method. They are used in `correlations.Rmd`
system("rm -r res")
system("mkdir res") # Create results folder
library(NMF)
library(ggfortify)
library(pheatmap)
signature      <- readLines("data.TCGA/EINAV_INTERFERON_SIGNATURE_IN_CANCER.txt") # Interferon signature
selected_genes <- "interferon_signature" 

pdf(file = paste0("res/dimreduction_", selected_genes, ".pdf"))
# All cancers with RNASeq2 data
cancer_RNASeq2 = c("ACC", "BLCA", "BRCA", "CESC", "CHOL", "COAD", "COADREAD", "DLBC", "ESCA", "GBM", "HNSC", "KICH", "KIPAN", "KIRC", "KIRP", "LIHC", "LUAD", "LUSC", "MESO", "OV", "PAAD", "PCPG", "PRAD", "READ", "SARC", "SKCM", "STAD", "TGCT", "THCA", "THYM", "UCEC", "UCS") # "GBMLGG", "LGG", 
data.type = "RNASeq2"; type = "" 
# Process each cancer
for (cancer_type in cancer_RNASeq2) {
  print(paste0("Processing cancer ", cancer_type))
  mtx <- load_data(disease = cancer_type, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
  # Prepare subset of expression data for a given signature
  mtx_selected <- t(log2(mtx$merged.dat[, signature]))
  colnames(mtx_selected) <- mtx$merged.dat$bcr
  # NMF 
  method <- "NMF"
  mtx_nmf <- nmf(t(mtx_selected), 3) # with three components
  pheatmap(mtx_selected[, order(mtx_nmf@fit@W[, 1])], cluster_cols = FALSE, treeheight_row = FALSE,   treeheight_col = FALSE, scale = "row", main = paste0(cancer_type, "_", method))
  mtx_reduced <- mtx_nmf@fit@W
  save(mtx_reduced, file = paste0("res/", cancer_type, "_", selected_genes, "_", method, ".Rda"))
  # PCA
  # https://cran.r-project.org/web/packages/ggfortify/vignettes/plot_pca.html
  method <- "PCA"
  mtx_pca <- prcomp(t(mtx_selected), scale. = TRUE, center = TRUE)
  pheatmap(mtx_selected[, order(mtx_pca$x[, 1])], cluster_cols = FALSE, treeheight_row = FALSE, treeheight_col = FALSE, scale = "row", main = paste0(cancer_type, "_", method))
  p <- autoplot(prcomp(t(mtx_selected), scale. = TRUE, center = TRUE), loadings = TRUE, loadings.label = TRUE) + ggtitle(paste0(cancer_type, "_", method))
  print(p)
  mtx_reduced <- mtx_pca$x
  save(mtx_reduced, file = paste0("res/", cancer_type, "_", selected_genes, "_", method, ".Rda"))
  # Factor analysis
  method <- "FA"
  mtx_fa <- factanal(t(mtx_selected), factors = 3, scores = "regression")
  pheatmap(mtx_selected[, order(mtx_fa$scores[, 1])], cluster_cols = FALSE, treeheight_row = FALSE, treeheight_col = FALSE, scale = "row", main = paste0(cancer_type, "_", method))
  p <- autoplot(mtx_fa,  loadings = TRUE, loadings.label = TRUE) + ggtitle(paste0(cancer_type, "_", method))
  print(p)
  mtx_reduced <- mtx_fa$scores
  save(mtx_reduced, file = paste0("res/", cancer_type, "_", selected_genes, "_", method, ".Rda"))
}
dev.off()
system("mv res res.genes.Analysis6")
```

# References