---
title: Differential expression analysis of TCGA data between clinical subgroups
author: "Mikhail Dozmorov"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: no
  html_document:
    theme: united
    toc: no
csl: styles.ref/genomebiology.csl
bibliography: data.TCGA/TCGA.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
# Set up the environment
library(knitr)
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=100, warnings=F, message=F, comment=NA, warning=F, results='as.is') #out.width=700, 
library(pander)
panderOptions('table.split.table', Inf)
set.seed(1)
library(dplyr)
options(stringsAsFactors = FALSE)
```

```{r results='hide'}
library(TCGA2STAT)
library(dplyr)
library(knitr)
library(ggplot2)
library(reshape2)
library(gridExtra)
library(limma)
library(openxlsx)
library(MDmisc)
library(org.Hs.eg.db)
# devtools::install_github("mdozmorov/enrichR")
library(enrichR)
```

```{r}
# A function to load TCGA data, from remote repository, or a local R object
load_data <- function(disease = cancer, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE) {
  FILE = paste0(data_dir, "/mtx_", disease, "_", data.type, "_", type, ".rda") # R object with data
  if (all(file.exists(FILE), !(force_reload))) {
    # If the data has been previously saved, load it
    load(file = FILE)
  } else {
    # If no saved data exists, get it from the remote source
    mtx <- getTCGA(disease = disease, data.type = data.type, type = type, clinical = TRUE)
    save(file = FILE, list = c("mtx")) # Save it
  }
  return(mtx)
}

# A function to get data overview
summarize_data <- function(mtx = mtx) {
  print(paste0("Dimensions of expression matrix, genex X patients: ", paste(dim(mtx$dat), collapse = " ")))
  print(paste0("Dimensions of clinical matrix, patients X parameters: ", paste(dim(mtx$clinical), collapse = " ")))
  print(paste0("Dimensions of merged matrix, patients X parameters + genes: ", paste(dim(mtx$merged.dat), collapse = " ")))
  print("Head of the merged matrix")
  print(mtx$merged.dat[1:5, 1:10])
  print("Head of the clinical matrix")
  print(mtx$clinical[1:5, 1:7])
  print("List of clinical values, and frequency of each variable: ")
  clin_vars <- apply(mtx$clinical, 2, function(x) length(table(x[ !(is.na(x) & x != "" )]))) %>% as.data.frame()
  # Filter clinical variables to have at least 2, but no more than 10 categories,
  # And they are not dates
  clin_vars <- clin_vars[ as.numeric(clin_vars$.) > 1 & as.numeric(clin_vars$.) < 10 & !grepl("years|days|date|vital", rownames(clin_vars), perl = TRUE) , , drop = FALSE]
  print(kable(clin_vars))
  return(rownames(clin_vars))
}
# A wrapper function to perform all functional enrichment analyses.
# Helper function to save non-empty results
save_res <- function(res, fileName = fileName, wb = wb, sheetName = "KEGG") {
  if (nrow(res) > 0) {
    openxlsx::addWorksheet(wb = wb, sheetName = sheetName)
    openxlsx::writeData(wb, res, sheet = sheetName)
    openxlsx::saveWorkbook(wb, fileName, overwrite = TRUE)
  }
}

# A wrapper to save the results
save_enrichr <- function(up.genes = up.genes, dn.genes = NULL, databases = "KEGG_2016", fdr.cutoff = 1, fileNameOut = NULL, wb = NULL) {
  print(paste("Running", databases, "analysis", sep = " "))
  if (is.null(dn.genes)) {
    res.kegg <- enrichGeneList(up.genes, databases = databases, fdr.cutoff = fdr.cutoff)
  } else {
    res.kegg <- enrichFullGeneList(up.genes, dn.genes, databases = databases, fdr.cutoff = fdr.cutoff)
  }
  
  res.kegg$pval <- formatC(res.kegg$pval, digits = 3, format = "e")
  res.kegg$qval <- formatC(res.kegg$qval, digits = 3, format = "e")
  if (!is.null(fileNameOut)) {
    if (nchar(databases) > 30) databases <- paste0(substr(databases, 1, 20), "_", substr(databases, nchar(databases) - 8, nchar(databases))) # If a database is longer that 30 characters, keep first 20 and last 10 characters
    save_res(res.kegg, fileNameOut, wb = wb, sheetName = databases)
  }
  # Pause for a few seconds
  pause_sec <- round(runif(1, min = 1, max = 10))
  Sys.sleep(pause_sec)
  return(res.kegg)
}
```

# Methods

```{r echo=TRUE}
# Cancer type
cancer = "LIHC" 
# Minimum number of samples in a subcategory to be considered
min_samples <- 15
# Functional category, from http://amp.pharm.mssm.edu/Enrichr/#stats. If NA, no enrichment analysis
enrichr_category <- NA # c("KEGG_2016", "Chromosome_Location",  "GO_Biological_Process_2017b", "GO_Molecular_Function_2017b") # ,

# Path where the downloaded data is stored
data_dir = "/Users/mdozmorov/Documents/Data/GenomeRunner/TCGAsurvival/data" # Mac

# General settings
useTPM    = FALSE # Whether or not to convert expression counts to TPM
# https://github.com/mksamur/RTCGAToolbox/issues/3
data.type = "RNASeq2" #  "CNA_SNP" # "CNV_SNP" # "RNASeq2"
type = "" 

# Selected clinical subcategory, will be run if not NA. Defined from summarize_data(mtx = mtx)
selected_clinical_subcategory <- "race" # Set to NA to run all clinical subcategories
selected_clinical_subgroup1   <- "black or african american" # Set both subgroups to NA
selected_clinical_subgroup2   <- "white"                     # To run all pairs of subgroups
  
# Differential expression cutoff
pval_cutoff   <- 0.74 # Adjusted P-value cutoff
max_num_genes <- 5000 # Maximum number of genes to feed into functional enrichment
full_degs_output <- TRUE # If TRUE, output differential expression statistics for all genes

# Enrichment cutoffs
fdr.cutoff     <- 0.1
top_X          <- 10 # How many top significant differentially expressed genes/pathways to output

# Filename to same the results
fileNameRes <- paste0("results/", cancer, "_", data.type, "_DEGs_clin_subcategories.xlsx")
```

```{r}
# Do TPM conversion, if needed
if (useTPM) {
  # Check if the TPMs have already been precalculated for a given cancer
  fileNameTPM <- paste0(data_dir, "/", cancer, "_TPM.Rda")
  if (!file.exists(fileNameTPM)) {
    source("calcTPM.R")
    load(file = "data/feature_length.Rda")
    common_genes <- intersect(colnames(expr), feature_length$Symbol) # Common gene symbols
    expr         <- expr[, c("AffyID", common_genes)] # Subset expression matrix
    feature_length <- feature_length[feature_length$Symbol %in% common_genes, ] # Subset feature length
    feature_length <- feature_length[match(colnames(expr)[ -1 ], feature_length$Symbol), ] # Match order
    all.equal(colnames(expr), c("AffyID", feature_length$Symbol)) # Should be true
    expr_tpm <- calcTPM(t(expr[, -1]), feature_length = feature_length) # Convert to TPM, takes time
    expr_tpm <- data.frame(AffyID = expr[, "AffyID"], t(expr_tpm), stringsAsFactors = FALSE)
    expr <- expr_tpm
    save(list = c("expr"), file = fileNameTPM) # Save the object
  } else {
    load(file = fileNameTPM)
  }
}
```

## Differential expression analysis

Samples in the selected cancer cohort were separated into clinical subgroups, e.g., within "race" clinical category pairs of subcategories, e.g., "black or african american" vs. "white" subgroups were compared for differentially expressed genes using `limma` v `r packageVersion("limma")` R package [@Ritchie:2015aa; @Smyth:2004aa]. P-values were corrected for multiple testing using False Discovery Rate (FDR) method [@Benjamini:1995aa]. Genes differentially expressed at unadjusted p-value cutoff `r pval_cutoff` were selected for '`r enrichr_category`' functional enrichment analysis.

Results are stored in the Excel file `r fileNameRes`. 

- Sheets are named by numbers
- One sheet contains differentially expressed genes, followed by sheets with functional analysis results 
- Each results output has a header explaining what the results are

- Legend for differentially expressed gene lists: "Gene" - gene annotations; "logFC" - log fold change; "AveExpr" - average expression, log2; "t" - t-statistics; "P.Val"/"adj.P.Val" - non-/FDR-adjusted p-value, "B" - another statistics.

- Legend for functional enrichment analysis: "database" - source of functional annotations, "category" - name of functional annotation,  "pval" - unadjusted enrichment p-value,  "qval" - FDR-adjusted p-value,  "genes" - comma-separated differentially expressed genes enriched in a corresponding functional category.

```{r}
# Create results folder
system("mkdir -p results") 
# Initialize results file
unlink(fileNameRes)
wb <- openxlsx::createWorkbook(fileNameRes) # openxlsx::loadWorkbook(fileName)
sheetCount <- 0 # Simple sheet name placeholder

cancer_type <- cancer
print(paste0("Processing cancer ", cancer_type))
mtx <- load_data(disease = cancer_type, data.type = data.type, type = type, data_dir = data_dir, force_reload = FALSE)
clinical_annotations <- summarize_data(mtx = mtx)
# Prepare expression data
expr <- mtx$merged.dat[ , 4:ncol(mtx$merged.dat)] %>% as.matrix
# Filter out low expressed genes
# Should be more than 90% of non-zero values
# ff <- genefilter::pOverA(p = 0.9, A = 0, na.rm = TRUE) 
# expr <- expr[, apply(expr, 2, ff)] 
expr <- data.frame(AffyID = mtx$merged.dat$bcr, expr, stringsAsFactors = FALSE)
# Prepare clinical data
clin <- mtx$merged.dat[, 1:3]
colnames(clin)[1] <- "AffyID"
# Full clinical information
clin_full <- mtx$clinical[rownames(mtx$clinical) %in% clin$AffyID, ]
clin_full <- clin_full[match(expr$AffyID, rownames(clin_full)), ]
all.equal(rownames(clin_full), expr$AffyID)
# Save clinical data
# wb1 <- openxlsx::createWorkbook(paste0("results/", cancer, "_clinical.xlsx")) # openxlsx::loadWorkbook(fileName)
# save_res(data.frame(Patients = rownames(clin_full), clin_full), fileName = paste0("results/", cancer, "_clinical.xlsx"), wb = wb1, sheetName = paste0(cancer, "_clinical"))


# Override clinical subcategories, if non-empty
if (!is.na(selected_clinical_subcategory) | selected_clinical_subcategory != "") {
  clinical_annotations <- selected_clinical_subcategory
  table(clin_full[, clinical_annotations], useNA = "no") # Check subgroups
}

# For each clinical annotation
for (annotation in clinical_annotations) { 
  # Get the number of patients per category in the current annotation
  annotations <- table(clin_full[, annotation], useNA = "no") 
  # Remove annotations less than a pre-defined number of samples
  annotations <- annotations[annotations >= min_samples]
  # Get combinations of all pairs of clinical subgroups
  annotations_combn <- combn(names(annotations), 2)
  # If custom subgroups selected, subset the combination matrix
  if (!(is.na(selected_clinical_subgroup1) & is.na(selected_clinical_subgroup2))) {
    annotations_combn <- annotations_combn[, grepl(selected_clinical_subgroup1, annotations_combn[1, ]) &
                                             grepl(selected_clinical_subgroup2, annotations_combn[2, ]), drop = FALSE]
  }
  # Go through each combination
  for (i in 1:ncol(annotations_combn)) {
    group1 <- annotations_combn[1, i] # First group
    group2 <- annotations_combn[2, i] # second group
    # Samples for second vs. first group comparison
    ind_up <- which(rownames(clin_full) %in% rownames(clin_full[ clin_full[, annotation] %in% group2 , ]))     
    ind_lo <- which(rownames(clin_full) %in% rownames(clin_full[ clin_full[, annotation] %in% group1 , ]))
    # For differential analysis, create group labels
    group <- vector(mode = "numeric", length = nrow(expr)) # Empty bector
    group[ind_up] <- 1 # Assign numeric groups
    group[ind_lo] <- 2
    # table(group) # How many patients we have
    expr_subset <- expr[group != 0, ] # Remove those that are not in quantiles
    group <- group[ group != 0 ]
    # Reshape expr_subsetession matrix
    expr_subset <- (t(expr_subset))
    colnames(expr_subset) <- expr_subset[1, ]
    expr_subset <- expr_subset[-1, ]
    class(expr_subset) <- "numeric"
    # Convert to the proper numerical scale for limma
    if (grepl("RNA", data.type)) {
      expr_subset <- voom(expr_subset)$E
    }
    if (grepl("CN", data.type)) {
      expr_subset <- ((1 + expr_subset) / 2) * 10
    }
    # boxplot(expr_subset)
    
    # Limma
    design <- model.matrix(~0 + factor(group))
    colnames(design) <- c("up", "lo")
    fit <- lmFit(expr_subset, design)
    contrast.matrix <- makeContrasts(up-lo, levels = design)
    fit2 <- contrasts.fit(fit, contrast.matrix)
    fit2 <- eBayes(fit2)
    degs <- topTable(fit2, coef = 1, number = Inf, p.value = pval_cutoff)
    
    # Check if the list of DEGs is non-empty, proceed with enrichment analysis
    if (nrow(degs) > 0) {
      # Subset the number of DEGs for KEGG analysis to the pre-defined maximum
      if (nrow(degs) > max_num_genes) {
        degs_subset <- degs[1:max_num_genes, ]
      } else {
        degs_subset <- degs
      }

      # Save DEGs, with header
      sheetCount <- sheetCount + 1 # Increase sheet count
      if (full_degs_output) { # Save the full DEG list?
        degs_to_save <- topTable(fit2, coef = 1, number = Inf)
        degs_to_save <- data.frame(Gene = rownames(degs_to_save), degs_to_save) # Append gene names to the DEG list
      } else {
        degs_to_save <- data.frame(Gene = rownames(degs), degs) #Append gene names to the DEG list
      }
      # Make header    
      header <- data.frame(matrix("", ncol = ncol(degs_to_save), nrow = 1, dimnames = list(" ", colnames(degs_to_save)))) # Header one-row data frame, column names matching DEGs
      header[1, 1] <- paste0(nrow(degs), " DEGs in ", cancer, ", between groups \'", annotation, ":", group2, "\' vs. \'", annotation, ":", group1, "\'") # Construct header
      save_res(rbind(header, degs_to_save), fileName = fileNameRes, wb = wb, sheetName = as.character(sheetCount))
      
      # Get list of up- and downregulated genes
      up.genes <- sort(unique(rownames(degs_subset)[ degs_subset$t > 0 ]))
      dn.genes <- sort(unique(rownames(degs_subset)[ degs_subset$t < 0 ]))
      # Run functional enrichment, if not NA
      if (!is.na(enrichr_category)) {
        for (j in enrichr_category) {
          res.kegg <- save_enrichr(up.genes = unique(c(up.genes, dn.genes)), databases = j, fdr.cutoff = fdr.cutoff)
          # Save the results, if non-empty
          if (nrow(res.kegg) > 0) {
            sheetCount <- sheetCount + 1 # Increase sheet count
            header <- data.frame(matrix("", ncol = ncol(res.kegg), nrow = 1, dimnames = list(" ", colnames(res.kegg)))) # Header one-row data frame, column names matching DEGs
            header[1, 1] <- paste0(nrow(res.kegg), " enrichments in ", j, " in ", cancer, ", between groups \'", annotation, ":", group2, "\' vs. \'", annotation, ":", group1, "\'") # Construct header
            save_res(rbind(header, res.kegg), fileName = fileNameRes, wb = wb, sheetName = as.character(sheetCount))
          }
        }
      }
    }
  }
}
```


# References
